Abstractions for Multi-Sorted Substitutions

### 2024-03-14 paper/paper.tex
Update.

  A \ACode{\ACon{CTraversal}} provides a generic \ACode{\AField{fusion}} lemma
  that works for any composition described by a \ACode{\ACon{CKit}}:
  \LibCode\KCTraversal

@@

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \end{itemize}
  We define \ACode{\ACon{CKit}} instances for each of the four
  combinations between renamings and substitutions:
  \LibCode\KCKitInstances
  Interestingly, the \ACode{\ADef{Cₛ}} instance has another
  \ACode{\ADef{CKit}} parameter.
  This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
  which yields the \ACode{\ADef{CKit}} between substitutions
  and renamings, and then use this resulting \ACode{\ADef{CKit}}
  as an argument for the case with two substitutions.
  As shown below, Agda's instance resolution is up for the task:

@@

  In the context of multi-sorted syntax, the notion of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}}, and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  The following structure is used to teach our framework about types:
  \LibCode\KTypeSorts
  For System F the instantiation is
  \AppCode\FTypes
  There are two things to discuss:
  \begin{enumerate}
  \item
    The \ACode{\AField{↑ᵗ}} function maps a sort of arbitrary sort type,
    to a sort of a potentially different sort type, which is expressed
    by the use of an existential.
    For System F we require this generality, as the sort
    \ACode{\ACon{𝕥}} can have variables, whereas its corresponding type
    sort \ACode{\ACon{𝕜}} cannot.
  \item
    Some sorts, like \ACode{\ACon{𝕜}}, do not have corresponding type
    sorts, but we still need to provide one, as \ACode{\AField{↑ᵗ}} is a
    total function.
    For such sorts, we can simply use arbitrary sort types, as the
    formalization will never use them.
  \end{enumerate}
  To hide the existential, we define \ACode{S \ACon{∶⊢} s}, which represents
  the type for a term \ACode{S \ACon{⊢} s}.
  \LibCode\KTypes

@@

  \noindent
  \begin{minipage}[t]{1.0\linewidth}
    \begin{minipage}[t]{0.55\linewidth}
      \LibCode\KContextLookupI
    \end{minipage}
    \begin{minipage}[t]{0.45\linewidth}
      \LibCode\KContextLookupII
    \end{minipage}
  \end{minipage}
  % \LibCode\KContextLookup

>>>>

  A \ACode{\ACon{CTraversal}} provides a \ACode{\AField{fusion}} lemma
  that works for the composition of any \ACode{\ACon{CKit}}:
  \LibCode\KCTraversal

@@

    % A weakening followed by a singleton substitution cancel each other out:
    % \LibCode\KWeakenCancelsSingle
    % % \LibCode\KWeakenCancelsSingleProof
    % By applying \ACode{\AField{fusion}} and \ACode{\AField{⋯-id}} to
    % this lemma, we get the same statement for multiple map applications:
    % \LibCode\KWeakenCancelsSingleTraverse
    % % \LibCode\KWeakenCancelsSingleTraverseProof
  \item 
    A singleton map can be swapped with any map \ACode{ϕ}:
    \LibCode\KDistLiftSingleTraverse
    % A singleton map can be swapped with any map \ACode{ϕ}:
    % \LibCode\KDistLiftSingle
    % % \LibCode\KDistLiftSingleProof
    % By applying \ACode{\AField{fusion}} to this lemma, we get the
    % same statement for multiple map applications:
    % \LibCode\KDistLiftSingleTraverse
    % % \LibCode\KDistLiftSingleTraverseProof
  \end{itemize}
  Similarly, as it was the case for the \ACode{\ADef{Kit}} and \ACode{\ADef{Traversal}}
  structures, the idea is that we instantiate the \ACode{\ADef{CTraversal}} for
  our object language, and in return the framework defines the concrete \ACode{\ADef{CKit}}
  instances for us:
  \LibCode\KCKitInstances
  Interestingly, the \ACode{\ADef{Cₛ}} instance has another
  \ACode{\ADef{CKit}} parameter.
  This allows us to first use \ACode{\ADef{Cᵣ}} as an argument,
  which yields the \ACode{\ADef{CKit}} between substitutions
  and renamings, and then use this resulting \ACode{\ADef{CKit}}
  as an argument for the case with two substitutions.
  As shown below, Agda's instance resolution is up for the task:

@@

  In the context of multi-sorted syntax, the notion of a type can be
  described as a mapping between sorts.
  For System F, the expression sort \ACode{\ACon{𝕖}} maps to the type sort
  \ACode{\ACon{𝕥}}, and the sort \ACode{\ACon{𝕥}} maps to the kind sort
  \ACode{\ACon{𝕜}}.
  The following structure is used to teach our framework about types:
  \LibCode\KTypeSorts
  For System F, the instantiation is
  \AppCode\FTypes
  There are two things to discuss:
  \begin{enumerate}
  \item
    The \ACode{\AField{↑ᵗ}} function maps a sort of arbitrary sort type,
    to a sort of a potentially different sort type, which is expressed
    by the use of an existential.
    For System F we require this generality, as the sort
    \ACode{\ACon{𝕥}} can have variables, whereas its corresponding type
    sort \ACode{\ACon{𝕜}} cannot.
  \item
    Some sorts, like \ACode{\ACon{𝕜}}, do not have corresponding type
    sorts, but we still need to provide one, as \ACode{\AField{↑ᵗ}} is a
    total function.
    For such sorts, we can simply use arbitrary sort types, as the
    formalization will have no typing rules that use them.
  \end{enumerate}
  To hide the existential, we define \ACode{S \ACon{∶⊢} s}, which represents
  the type for a term \ACode{S \ACon{⊢} s}.
  \LibCode\KTypes

@@

  \noindent
  \begin{minipage}[t]{1.0\linewidth}
    \begin{minipage}[t]{0.45\linewidth}
      \LibCode\KContextLookupII
    \end{minipage}
    \begin{minipage}[t]{0.55\linewidth}
      \LibCode\KContextLookupI
    \end{minipage}
  \end{minipage}
  % \LibCode\KContextLookup
